pdp: An R Package for Constructing Partial Dependence Plots

### 2016-12-16 manuscript/reviews/response.Rmd

---
title: "Response to Comments"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
  word_document: default
---

Thanks to the reviewers and editor for the well thought out comments. I believe that they have greatly imporved, not only the manuscript, but the package it describes as well. I hope that my responses are at least satisfactory. Please find below a copy of the reviewers comments along with my responses.

## ----------Reviewer 1----------

#### Comment
Bug: A graph is produced as expected if the code below is cut-and-pasted at the command line .  However, if the code is put into a file (say `foo.R`) and `source("foo.R")` is executed at the command line, no graph is produced.  (But note that if `plot=TRUE` is added to the call to `partial()`, then a plot is produced as expected.)
```{r, eval=FALSE}
library(randomForest)
data(airquality)
ozone.rf <- randomForest(Ozone ~ ., data=airquality, na.action=na.omit)
library(pdp)
pd <- partial(ozone.rf, pred.var="Temp")
plotPartial(pd)
```
#### Response
This is simply a side effect of using `lattice` to produce the plots. If sourcing a script, then the plot needs to be explicitly printed (e.g., `print(plotPartial(pd))`). This has been explained in both the documentation and the manuscript in a **Note:** on 3.

#### Comment
I couldn't get `partial` to work with `gbm` (version 2.1.1) models (although I didn't try very hard).  Perhaps this is because the `gbm` maintainer changed the `gbm` interface in a non-backcompatible way. The following (where `gbm.mod` is a model built using the gbm package)
```{r, eval=FALSE}
partial(gbm.mod, pred.var="age")
```
gives
```{r, eval=FALSE}
Error in paste("Using", n.trees, "trees...\n") :
argument "n.trees" is missing, with no default
```
I also tried
```{r, eval=FALSE}
partial(gbm.mod, pred.var="age", n.trees=100)
```
but that gives
```{r, eval=FALSE}
Error in .fun(piece, ...) : unused argument (n.trees = 100)
```
#### Response
The `gbm` maintainer did contact me a while back and this has been fixed in `pdp` (version 0.3.0). The user will still have to supply `n.trees` in the call to `partial`. The test suite is being expanded, as suggested in a later comment, to help catch more errors like this.

#### Comment
Standard partial matching isn't implemented (e.g., `partial(..., type="r")`).
#### Response
Partial matching is now implemented, but only for the `type` argument. Thanks for the suggestion.

#### Comment
The following code incorrectly gives an error message
```
"Error in names(pd_df) <- c(pred.var, "y") :
  'names' attribute [2] must be the same length as the vector [1]"
```
It seems to work ok if you change the variable names.
```{r, eval=FALSE}
library(pdp)
data <- data.frame(V1=1:10, V2=1:10)
mod <- lm(V2~V1, data=data)
partial(mod, pred.var="V1", plot=TRUE) # error
```
#### Response
This has been fixed in `pdp` (version 0.3.0). This was simply a side effect of `partial`'s internal use of `plyr::adply` which uses default variable names (e.g., `V1`, `V2`, etc.). This caused the column `V1` in the output data frame to be overwritten, producing the error. 
```{r, eval=TRUE}
library(pdp)
data <- data.frame(V1=1:10, V2=1:10)
mod <- lm(V2~V1, data=data)
partial(mod, pred.var="V1") # no more error
```

@@

#### Comment
I think you should mention in the documentation something about the assumptions you make about accessing the model data. The following (somewhat contrived) example fails with
`"Error in train[[x]] : subscript out of bounds":`
```{r, eval=FALSE}
data(trees)
foo <- function(data)
{
    lm(Volume~., data=data)
}
mod <- foo(trees)
library(pdp)
partial(mod, pred.var="Girth", plot=TRUE) # fails
```
and the following silently gives misleading results:

```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var="Girth", plot=TRUE) # ok
trees <- trees[1:2,]
partial(mod, pred.var="Girth", plot=TRUE) # wrong xlim
```
I'm not necessarily saying that you need to resolve the above two
issues, but at least mention in the documentation that this kind of
thing can happen, and how to avoid it (basically call partial in the
same environment used to build the model, and don't change any of the
data used to build the model).
#### Response
Aggred! Unfortunately, some functions just don't store a copy of the training data. I have expanded the code to work harder, especially for `S4` methods, to inform the user to supply the training data when necessary. The documentation for `partial` has also been expanded to discuss this.

#### Comment
The following code gives an obscure error message:
```
"Error in seq.default(from = min(train[[x]], na.rm = TRUE),
   to = max(train[[x]],  :
  'from' cannot be NA, NaN or infinite"
```
Some better handholding for the user may be helpful here.
```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var="nosuchvariable", plot=TRUE) # obscure err msg
```
#### Response
This has been fixed in `pdp` (version 0.3.0) by adding an informative error message listing which variables are the cause of the issue.
```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var=c("Girth", "foo", "bar"), plot=TRUE)
```
```
Error in partial.default(mod, pred.var = c("Girth", "foo", "bar"), plot = TRUE) : foo, bar not found in the training data.
```

#### Comment
On page 2 of the paper it says "The columns of the data frame are labeled in the same order as the features supplied to `pred.var`, and the last column is always labeled `y`". What happens if `y` is one of the predictors?
#### Response
The last column is now labeled `yhat`; which is less likely to be a predictor name. When one of the predictors is labeled `"y"` `randomForest::partialPlot` fails while `gbm::plot.gbm` does not. I think regardless of what the default column name is, there is always the possibility of this issue. This has been filed as an issue on the `pdp` GitHib page :

#### Comment
Hastie et al. in the book cited in the paper say "Although such a collection [of partial dependence plots] can seldom provide a comprehensive depiction of the approximation, it can often produce helpful clues".  Therefore in in the abstract for the paper, to avoid overselling consider changing "relationship between the outcome and predictors of interest can be easily understood" to "relationship between the outcome and predictors of interest can be more easily understood".
#### Response
Fixed. Thanks for the suggestion.

#### Comment
On page 3 in the PDF of the paper, add a newline to the error message in the paper:
```
Error: The training data could not be extracted from object. Please
supply the raw training data using the ...
```
#### Response
Fixed. Thanks for pointing that out.

#### Comment
Is there any reason the package is called `pdp` but the main function is called `partial`?  Consider giving the function the same name as the package, so e.g., the following works (to help the user get started)
```{r, eval=FALSE}
library(pdp)
?pdp # fails
```
#### Response
The package was originally called `partial`, but at the time of its original submission this conflicted with a CRAN package of the same name (but with mixed lowercase/uppercase letetrs). So, I decided to change the name of the package to `pdp`; besides, I think `partial` and `plotPartial` work well together as function names. However, I did add a help page describing the package, main functions, etc. So, `?pdp` no longer fails. I am, however, not opposed to creating an alias to `partial` called `pdp` in the future.

#### Comment
`partial` should return the same value regardless of whether or not the `plot` argument is used?
#### Response
I dont think so, and this is the same behavior for the PDP functions in `randomForest` and `gbm` (not that `pdp` needs to immitate the behavior of these packages PDP functions). However, I did add the partial dependence data as an attribute to the returned `"trellis"` object (i.e., plot) so that the user can extract the data easily from that if necessary (e.g., if they do not like the plot and don't want to re-run partial). For example,
```{r}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
pd.Girth <- partial(mod, pred.var = "Girth")  # data only
pdp.Girth <- partial(mod, pred.var = "Girth", plot = TRUE)  # plot
identical(pd.Girth, attr(pdp.Girth, "partial.data"))  # should be identical
```
Also, the `lattice` function `trellis.last.object` is now exported when `pdp` is loaded. This will be useful when the user does not store the resulting plot in the call to `partial` but needs to retrieve the outout data to avoid re-running `partial` when computation time is a concern. For example,
```{r}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
partial(mod, pred.var = "Girth", plot = TRUE)
pdp.Girth <- lattice::trellis.last.object()
pd.Girth <- attr(pdp.Girth, "partial.data")
head(pd.Girth)
```
Thanks for the suggestion!

#### Comment
A call to say `par(mfrow = c(2, 2))` before calling `plotPartial` is ignored by `plotPartial`.  This is a pity because it would allow multiple plots to be put on one page like this:
```{r, eval=FALSE}
par(mfrow=c(2,2))
partial(ozone.rf, pred.var = "Temp", plot = TRUE)
partial(ozone.rf, pred.var = "Wind", plot = TRUE)
```
#### Response
The command `par(mfrow = c(2, 2))` does not work with `"trellice"` objects, such as those produced by `lattice` which `partial` relies on for producing its plots. Other methods are available. For example, using the `gridExtra` package:
```{r, eval=TRUE}
library(randomForest)
data(airquality)
ozone.rf <- randomForest(Ozone ~ ., data=airquality, na.action=na.omit)
library(pdp)
pdp1 <- partial(ozone.rf, pred.var = "Temp", plot = TRUE)
pdp2 <- partial(ozone.rf, pred.var = "Wind", plot = TRUE)
gridExtra::grid.arrange(pdp1, pdp2, ncol = 2)
```
For convenience, `pdp` now imports the `grid.extra` function and this has been incorporated into the manuscripts examples where appropriate.

However, since `partial` returns a data frame by default, using `par` would work in the following case
```{r, eval=FALSE}
par(mfrow=c(2,2))
plot(partial(ozone.rf, pred.var = "Temp"), type = "l")
plot(partial(ozone.rf, pred.var = "Wind"), type = "l")
```

#### Comment
It would be nice if the functions automatically plotted all the (important) variables in a grid of plots, without forcing the user to explicitly specify them.
#### Response 
I agree, this would be a nice feature, but the problem I see is that not all models emit a "variable importance". The `caret` package has a come a long way in this regard. At this point, I think it is simpler to have the user select which variables they would like to see plotted (it forces them to explore the results). I am not opposed to including this feature in the future. I added added a possible solution along with a feature request: https://github.com/bgreenwell/pdp/issues/19.

#### Comment
It would be nice if the functions automatically got the data from the model used to build the model in more cases without forcing the user to explicitly specify the data.  Since XGBoost seems now quite popular, perhaps the authors of that package could be persuaded to make the original model data available for functions like partial (currently XGBoost uses a custom object `xgb.DMatrix` which is currently write-only -- one can't access the data stored in `xgb.DMatrix`).
#### Response
This is also true for the fitting functions in other package (e.g., `kernlab::ksvm`). As time goes on, the coverage will certainly be inncreased to make it easier on the user. For example, `partial` now works with `caret` which, by default, stores the taining data in a component called `trainingData`. I suspect that the package maintainers for `xgboost`  would be hesitent in storing a copy of the training data, especially since the package is often used for modeling large data sets; they probably do not want to store a copy of the data with the fitted model object for memory reasons.

### 2016-12-16 manuscript/greenwell.tex
tweaks based on review comments

\citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data. One of their goals was to find a housing value equation using data on median housing values from 506 census tracts in the suburbs of Boston from the 1970 census. The prediction equation from one of there regression analyses is reproduced in Equation~\eqref{eqn:boston}. \citet{harrison-1978-hedonic} took great care and time to find this model which obtained an $R^2$ of roughly 0.81. Part of the payoff for there efforts was a reasonable fitting model with an interpretable equation. For example, the equation tells us that an increase in median home value is associated with an increase in the average number of rooms per dwelling.
\begin{equation}
\label{eqn:boston}
\begin{split}
\widehat{\log\left(MV\right)} &= 9.76 + 0.0063 RM^2 + 8.98\times10^{-5} AGE - 0.19\log\left(DIS\right) + 0.096\log\left(RAD\right) \\
  &\quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
  &\quad - 0.012 CRIM + 8.03\time10^{-5} ZN + 2.41\times10^{-4} INDUS + 0.088 CHAS - 0.0064 NOX^2
\end{split}
\end{equation}

Many modern nonparametric models, however, can fit the data automatically in seconds---typically with higher accuracy. The downfall, however, is some loss of interpretation; for example, there is usually no prediction formula like in Equation~\eqref{eqn:boston}. Fortunately, these methods can still provide insight into the data, but it is not in the form of simple equations. For example, many modern methods, like boosting and random forests, can assign importance scores to all of the predictors.

@@

\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics. \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}. \pkg{grid} graphics behave a little differently than traditional R graphics, and two points are worth making (see \code{?lattice} for more details):
\begin{enumerate}
  \item \pkg{lattice} functions return a \code{"trellis"} object, but do not display it; the \code{print} method produces the actual display. However, due to R's automatic printing rule, the result is automatically printed when using these functions in the command line. If \code{plotPartial} is called inside of \code{source} or inside a loop (e.g., \code{for} or \code{while}), an explicit \code{print} statement is required to display the resulting graph.
  \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots. Instead, \pkg{lattice} provides its own \code{trellis.par.set} function for modifying graphical parameters.
\end{enumerate}
A consequence of the second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots. Simple solutions are available in packages \CRANpkg{latticeExtra} \citep{latticeExtra-pkg} and \CRANpkg{gridExtra} \citep{gridExtra-pkg}. For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{pdp} plots on a single page.

@@

The \code{partial} function also supports object of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}. This means that \code{partial} can be used with any classification or regression model that has been fit using \pkg{caret}'s \code{train} function; see \url{http://topepo.github.io/caret/available-models.html} for a current list of models supported by \pkg{caret}. An example is given in Section~\ref{sec:xgboost}.

@@

Since \code{train} always stores a copy of the training data as part of the \code{"train"} object in a component called \code{trainingData} (e.g., \code{boston.xgb.tune\$trainingData}) there is no need to supply the training data in the call to \code{partial}. However, this is not the case when using the \pkg{xgboost} package directly. For example, the following code fits the same model using the \code{xgboost} function.
\begin{example}
set.seed(203)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 2000, max_depth = 3, eta = 0.01)
\end{example}
To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown). (Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
\begin{example}
partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
        train = subset(boston, select = -cmedv))
\end{example}

>>>>

\citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data. One of their goals was to find a housing value equation using data on median housing values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see Table IV of \citet{harrison-1978-hedonic} for a description of each variable. The data violate many classical assumptions like linearity, normality, and constant variance. Nonetheless, \citeauthor{harrison-1978-hedonic}---using a combination of transformations, significance testing, and grid searches---were able to find a reasonable fitting model ($R^2 = 0.81$). Part of the payoff for there efforts was an interpretable prediction equation which is reproduced in Equation~\eqref{eqn:boston}.
\begin{equation}
\label{eqn:boston}
\begin{split}
\widehat{\log\left(MV\right)} &= 9.76 + 0.0063 RM^2 + 8.98\times10^{-5} AGE - 0.19\log\left(DIS\right) + 0.096\log\left(RAD\right) \\
  & \quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
  & \quad - 0.012 CRIM + 8.03\time10^{-5} ZN + 2.41\times10^{-4}  INDUS + 0.088 CHAS - 0.0064 NOX^2
\end{split}
\end{equation}

Nowadays, supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy. (We will revisit the Boston housing data in Section~\ref{sec:boston}) The downfall, however, is some loss of interpretation; for example, there is usually no simple prediction formula like in Equation~\eqref{eqn:boston}. Fortunately, these methods can still provide insight into the data, but it is not in the form of simple equations. For example, many modern methods, like boosting and random forests, can assign importance scores to all of the predictors in the training data.

@@

\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics. \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}. \pkg{grid} graphics behave a little differently than traditional R graphics, and two points are worth making (see \code{?lattice} for more details):
\begin{enumerate}
  \item \pkg{lattice} functions return a \code{"trellis"} object, but do not display it; the \code{print} method produces the actual display. However, due to R's automatic printing rule, the result is automatically printed when using these functions in the command line. If \code{plotPartial} is called inside of \code{source} or inside a loop (e.g., \code{for} or \code{while}), an explicit \code{print} statement is required to display the resulting graph; hence, the same is true when using \code{partial} with \code{plot = TRUE}.
  \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots. Instead, \pkg{lattice} provides its own \code{trellis.par.set} function for modifying graphical parameters.
\end{enumerate}
A consequence of the second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots. Simple solutions are available in packages \CRANpkg{latticeExtra} \citep{latticeExtra-pkg} and \CRANpkg{gridExtra} \citep{gridExtra-pkg}. For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{pdp} plots on a single page.

@@

The \code{partial} function also supports objects of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}. This means that \code{partial} can be used with any classification or regression model that has been fit using \pkg{caret}'s \code{train} function; see \url{http://topepo.github.io/caret/available-models.html} for a current list of models supported by \pkg{caret}. An example is given in Section~\ref{sec:xgboost}.

@@

%Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object in a component called \code{trainingData} (e.g., \code{boston.xgb.tune\$trainingData}) there is no need to supply the training data in the call to \code{partial}. However, this is not the case when using the \pkg{xgboost} package directly. For example, the following code fits the same model using the \code{xgboost} function.
Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object there is no need to supply it in the call to \code{partial}. However, this is not the case when using the \pkg{xgboost} package directly. For example, the following code fits the same model using the \code{xgboost} function.
\begin{example}
set.seed(203)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 2000, max_depth = 3, eta = 0.01)
\end{example}
To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown). (Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
\begin{example}
partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
        train = subset(boston, select = -cmedv))
\end{example}
